---
title: "1 - gene vs metab native laptop"
output: html_document
date: "2026-03-23"
---

# load packages and files

```{r}
library(data.table)
library(tidyverse)
library(tibble)
library(Hmisc)
library(dplyr)
library(tidyr)

metadata <- read.csv("~/Documents/gutmicrobiome_pharmacology/hmp2_metadata_2018-08-20.csv")

gmetab_all <- read.csv("~/Documents/gutmicrobiome_pharmacology/metabolite_labels_updated_RL.csv")

#metabolites and transcriptomic files
metab <- read.csv("~/Documents/gutmicrobiome_pharmacology/METABOLOMICS w metadata updated.csv")
metab1 <- metab[,-1]

hxtx <- read.csv("~/Documents/gutmicrobiome_pharmacology/HOST TRANSCRIPTOMICS host_tx_counts FILTERED less zeros.csv")
hxtx1 <- hxtx[,-1]
```

# clean metadata

```{r}
temp <- metadata %>% 
  filter(data_type == "host_transcriptomics" | data_type == "metabolomics") %>%
  filter(site_name == "Cedars-Sinai" | site_name == "MGH") %>%
  mutate(data_type = as.factor(data_type)) %>%
  mutate(biopsy_location = as.factor(biopsy_location), 
         is_inflamed = as.factor(is_inflamed),
         diagnosis = as.factor(diagnosis),
         sex = as.factor(sex))

metadata_hxtx <- filter(temp, data_type =="host_transcriptomics")
metadata_metab <- filter(temp, data_type =="metabolomics") 
```

# RECTUM CORR, INFLAMED STATUS YES

## clean up host transcriptomics

```{r}
metadata_hxtx_rectum_yes <- metadata_hxtx %>%
  filter(biopsy_location == "Rectum",
         is_inflamed == "Yes",
         IntervalName == "Screening Colonoscopy")

# Match the External.IDs present in metadata_hxtx_rectum_yes
hxtx_location8 <- hxtx1 %>%
  dplyr::select(names(hxtx1)[1], all_of(metadata_hxtx_rectum_yes$External.ID))
```

## clean up metabolomics

```{r}
metadata_metab_weeks <- metadata_metab %>%
  filter(diagnosis %in% c("UC", "nonIBD", "CD")) %>%
  filter(IntervalName == "Stool Collection #1")

# Match the External.IDs present in metadata_metab_weeks
metab_location8 <- metab1 %>%
  dplyr::select(names(metab1)[1], all_of(metadata_metab_weeks$External.ID))

#REMOVING ROW 72 IN metab_location1 = metabolite C18n_TF7 = furosemide, as it has MAJORITY (162/164) zeros
metab_location8 <- metab_location8[-72, ]
```

## replacing External.ID to Participant.ID

```{r}
matching <- metadata_hxtx_rectum_yes %>%
  dplyr::select(External.ID, Participant.ID)
setnames(hxtx_location8, matching$External.ID, matching$Participant.ID)

matching2 <- metadata_metab_weeks %>%
  dplyr::select(External.ID, Participant.ID)
setnames(metab_location8, matching2$External.ID, matching2$Participant.ID)
```

## set up dataframes for correlations

```{r}
#transpose metab_location1
t_metab_location8 = as.data.frame(x = t(metab_location8), stringsAsFactors = FALSE)
t_metab_location8 <- t(metab_location8)
colnames(t_metab_location8) <- t_metab_location8[1, ]
t_metab_location8_1 <- t_metab_location8[-1,]

#transpose hxtx_location1
t_hxtx_location8 <- t(hxtx_location8)
colnames(t_hxtx_location8) <- t_hxtx_location8[1, ]
t_hxtx_location8_1 <- t_hxtx_location8[-1,]

#convert into dataframes
t_hxtx_location8_1 <- as.data.frame(t_hxtx_location8_1)
t_metab_location8_1 <- as.data.frame(t_metab_location8_1)

#intersect with common row names
common_row_names <- intersect(rownames(t_hxtx_location8_1), rownames(t_metab_location8_1))
hxtx_location8_1_common <- t_hxtx_location8_1[common_row_names, ]
metab_location8_1_common <- t_metab_location8_1[common_row_names, ]

df3 <- mutate_all(hxtx_location8_1_common, function(x) as.numeric(as.character(x)))
df4 <- mutate_all(metab_location8_1_common, function(x) as.numeric(as.character(x)))

df3 <- tibble::rownames_to_column(df3, "patient_id")
df4 <- tibble::rownames_to_column(df4, "patient_id")
```

## rename metabolites in df4 with gmetab_all as map

```{r}
gmetab_all <- gmetab_all %>%
  mutate(Metabolite_Compound = paste(Metabolite, Compound, sep = "_"))
         
name_map <- setNames(gmetab_all$Metabolite_Compound, gmetab_all$Compound)
colnames(df4) <- ifelse(colnames(df4) %in% names(name_map),
                        name_map[colnames(df4)],
                        colnames(df4))
```

## save files to run in HPC if have to

```{r}
write_csv(df3, "~/Documents/gutmicrobiome_pharmacology/df3_gene.csv")
write_csv(df4, "~/Documents/gutmicrobiome_pharmacology/df4_metab.csv")
```

## correlate

```{r}
# --- 1. Extract numeric matrices (drop patient_id column) ---
genes_mat <- as.matrix(df3[, -1])   # 34 x ~10000
metab_mat <- as.matrix(df4[, -1])   # 34 x ~650

rownames(genes_mat) <- df3$patient_id
rownames(metab_mat) <- df4$patient_id

# --- 2. Rank-transform columns (= Spearman) ---
genes_ranked <- apply(genes_mat, 2, rank)
metab_ranked <- apply(metab_mat, 2, rank)

# --- 3. Compute ALL correlations at once via matrix multiply ---
# rcorr on combined ranked matrix, then extract the cross-block
combined <- cbind(genes_ranked, metab_ranked)
rc <- Hmisc::rcorr(combined, type = "pearson")  # pearson on ranks = spearman

n_genes <- ncol(genes_ranked)
n_metab <- ncol(metab_ranked)

rho_mat <- rc$r[1:n_genes, (n_genes+1):(n_genes+n_metab)]
p_mat   <- rc$P[1:n_genes, (n_genes+1):(n_genes+n_metab)]
n_mat   <- rc$n[1:n_genes, (n_genes+1):(n_genes+n_metab)]

# --- 4. Pivot to long format ---
rho_long <- as.data.frame(rho_mat) %>%
  mutate(x = rownames(rho_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "rho")

p_long <- as.data.frame(p_mat) %>%
  mutate(x = rownames(p_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "p")

n_long <- as.data.frame(n_mat) %>%
  mutate(x = rownames(n_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "n")

gmr_corr <- rho_long %>%
  left_join(p_long, by = c("x", "y")) %>%
  left_join(n_long, by = c("x", "y")) %>%
  mutate(message = ifelse(is.na(rho), "insufficient data", "ok")) %>%
  # --- 5. FDR corrections (matching coracle's q_x, q_y, q) ---
  group_by(x) %>%
  mutate(q_x = p.adjust(p, method = "BH")) %>%   # FDR per gene (across all its metabolites)
  ungroup() %>%
  group_by(y) %>%
  mutate(q_y = p.adjust(p, method = "BH")) %>%   # FDR per metabolite (across all its genes)
  ungroup() %>%
  mutate(q = p.adjust(p, method = "BH")) %>%      # global FDR across all 6.5M pairs
  dplyr::select(x, y, rho, p, n, message, q_x, q_y, q)
```

## save corr file

```{r}
write_csv(gmr_corr, "~/Documents/gutmicrobiome_pharmacology/correlation gene and metab rectum adult updated 3-26-2026.csv")
```

# Partial correlations for RECTUM

## Diagnosis

```{r}
#replace diagnosis with category 0 and 1 where 0 = nonIBD and 1 = UC or CD 
metadata_hxtx_rectum_yes_IBD <- metadata_hxtx_rectum_yes %>%
  mutate(diagnosis = case_when(
    diagnosis == "nonIBD" ~ "0",
    diagnosis == "UC" ~ "1",
    diagnosis == "CD" ~ "1",
  ))

mtd_hxtx_rectum_pcor <- metadata_hxtx_rectum_yes_IBD %>%
  dplyr::select(Participant.ID, diagnosis)
mtd_hxtx_rectum_pcor <- mtd_hxtx_rectum_pcor %>% 
  mutate(Participant.ID2 = Participant.ID)
mtd_hxtx_rectum_pcor <- mtd_hxtx_rectum_pcor %>%
  dplyr::select(Participant.ID, Participant.ID2, diagnosis)
mtd_hxtx_rectum_pcor <- data.frame(mtd_hxtx_rectum_pcor, row.names = 1)

#intersect with common row names
mtd_hxtx_rectum_pcor_common_names <- intersect(rownames(mtd_hxtx_rectum_pcor), rownames(hxtx_location8_1_common))
mtd_hxtx_rectum_pcor_common <- mtd_hxtx_rectum_pcor[mtd_hxtx_rectum_pcor_common_names, ]

df34_diag <- as.data.frame(mtd_hxtx_rectum_pcor_common)
df34_diag <- tibble::rownames_to_column(df34_diag, "patient_id")
df34_diag <- df34_diag[,-2]
df34_diag$diagnosis <- as.numeric(as.character(df34_diag$diagnosis))
```

### correlate

```{r}
# --- 1. Align all dataframes by patient_id ---
common_patients <- Reduce(intersect, list(df3$patient_id, 
                                           df4$patient_id, 
                                           df34_diag$patient_id))

genes_mat <- as.matrix(df3[match(common_patients, df3$patient_id), -1])
metab_mat <- as.matrix(df4[match(common_patients, df4$patient_id), -1])
rownames(genes_mat) <- common_patients
rownames(metab_mat) <- common_patients

# --- 2. Build covariate matrix Z ---
diag_aligned <- df34_diag[match(common_patients, df34_diag$patient_id), ]
Z <- model.matrix(~ diagnosis, data = diag_aligned)  # intercept + diagnosis

# --- 3. Rank-transform columns ---
genes_ranked <- apply(genes_mat, 2, rank)
metab_ranked <- apply(metab_mat, 2, rank)

# --- 4. Residualize ranked data against covariate ---
residualize <- function(mat, Z) {
  (diag(nrow(mat)) - Z %*% solve(t(Z) %*% Z) %*% t(Z)) %*% mat
}

genes_resid <- residualize(genes_ranked, Z)
metab_resid <- residualize(metab_ranked, Z)

# --- 5. Correlate residuals ---
combined <- cbind(genes_resid, metab_resid)
rc <- Hmisc::rcorr(combined, type = "pearson")

n_genes <- ncol(genes_resid)
n_metab <- ncol(metab_resid)

rho_mat <- rc$r[1:n_genes, (n_genes+1):(n_genes+n_metab)]
p_mat   <- rc$P[1:n_genes, (n_genes+1):(n_genes+n_metab)]
n_mat   <- rc$n[1:n_genes, (n_genes+1):(n_genes+n_metab)]

# --- 6. Pivot to long and add FDR ---
rho_long <- as.data.frame(rho_mat) %>%
  mutate(x = rownames(rho_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "rho")

p_long <- as.data.frame(p_mat) %>%
  mutate(x = rownames(p_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "p")

n_long <- as.data.frame(n_mat) %>%
  mutate(x = rownames(n_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "n")

pcorr_diag_gmr <- rho_long %>%
  left_join(p_long, by = c("x", "y")) %>%
  left_join(n_long, by = c("x", "y")) %>%
  mutate(z = "diagnosis",                                      # <-- z column
         message = ifelse(is.na(rho), "insufficient data", "ok")) %>%
  group_by(x) %>%
  mutate(q_x = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  group_by(y) %>%
  mutate(q_y = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  mutate(q = p.adjust(p, method = "BH")) %>%
  dplyr::select(x, y, z, rho, p, n, message, q_x, q_y, q)
```

### save pcorr file

```{r}
write_csv(pcorr_diag_gmr, "~/Documents/gutmicrobiome_pharmacology/partial correlation gene and metab rectum cf diagnosis adult updated 3-26-2026.csv")
```

## Sex

```{r}
#replace sex with category 0 and 1 where 0 = male and 1 = female
metadata_hxtx_rectum_yes_SEX <- metadata_hxtx_rectum_yes %>%
  mutate(sex = case_when(
    sex == "Male" ~ "0",
    sex == "Female" ~ "1",
  ))

mtd_hxtx_rectum_pcor2 <- metadata_hxtx_rectum_yes_SEX %>%
  dplyr::select(Participant.ID, sex)
mtd_hxtx_rectum_pcor2 <- mtd_hxtx_rectum_pcor2 %>% 
  mutate(Participant.ID2 = Participant.ID)
mtd_hxtx_rectum_pcor2 <- mtd_hxtx_rectum_pcor2 %>%
  dplyr::select(Participant.ID, Participant.ID2, sex)
mtd_hxtx_rectum_pcor2 <- data.frame(mtd_hxtx_rectum_pcor2, row.names = 1)

#intersect with common row names
mtd_hxtx_rectum_pcor_common_names2 <- intersect(rownames(mtd_hxtx_rectum_pcor2), rownames(hxtx_location8_1_common))
mtd_hxtx_rectum_pcor_common2 <- mtd_hxtx_rectum_pcor2[mtd_hxtx_rectum_pcor_common_names2, ]

df34_sex <- as.data.frame(mtd_hxtx_rectum_pcor_common2)
df34_sex <- tibble::rownames_to_column(df34_sex, "patient_id")
df34_sex <- df34_sex[,-2]
df34_sex$sex <- as.numeric(as.character(df34_sex$sex))
```

### correlate

```{r}
# --- 1. Align all dataframes by patient_id ---
common_patients <- Reduce(intersect, list(df3$patient_id, 
                                           df4$patient_id, 
                                           df34_sex$patient_id))

genes_mat <- as.matrix(df3[match(common_patients, df3$patient_id), -1])
metab_mat <- as.matrix(df4[match(common_patients, df4$patient_id), -1])
rownames(genes_mat) <- common_patients
rownames(metab_mat) <- common_patients

# --- 2. Build covariate matrix Z ---
sex_aligned <- df34_sex[match(common_patients, df34_sex$patient_id), ]
Z <- model.matrix(~ sex, data = sex_aligned)  # intercept + sex

# --- 3. Rank-transform columns ---
genes_ranked <- apply(genes_mat, 2, rank)
metab_ranked <- apply(metab_mat, 2, rank)

# --- 4. Residualize ranked data against covariate ---
residualize <- function(mat, Z) {
  (diag(nrow(mat)) - Z %*% solve(t(Z) %*% Z) %*% t(Z)) %*% mat
}

genes_resid <- residualize(genes_ranked, Z)
metab_resid <- residualize(metab_ranked, Z)

# --- 5. Correlate residuals ---
combined <- cbind(genes_resid, metab_resid)
rc <- Hmisc::rcorr(combined, type = "pearson")

n_genes <- ncol(genes_resid)
n_metab <- ncol(metab_resid)

rho_mat <- rc$r[1:n_genes, (n_genes+1):(n_genes+n_metab)]
p_mat   <- rc$P[1:n_genes, (n_genes+1):(n_genes+n_metab)]
n_mat   <- rc$n[1:n_genes, (n_genes+1):(n_genes+n_metab)]

# --- 6. Pivot to long and add FDR ---
rho_long <- as.data.frame(rho_mat) %>%
  mutate(x = rownames(rho_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "rho")

p_long <- as.data.frame(p_mat) %>%
  mutate(x = rownames(p_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "p")

n_long <- as.data.frame(n_mat) %>%
  mutate(x = rownames(n_mat)) %>%
  pivot_longer(-x, names_to = "y", values_to = "n")

pcorr_sex_gmr <- rho_long %>%
  left_join(p_long, by = c("x", "y")) %>%
  left_join(n_long, by = c("x", "y")) %>%
  mutate(z = "sex",                                      # <-- z column
         message = ifelse(is.na(rho), "insufficient data", "ok")) %>%
  group_by(x) %>%
  mutate(q_x = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  group_by(y) %>%
  mutate(q_y = p.adjust(p, method = "BH")) %>%
  ungroup() %>%
  mutate(q = p.adjust(p, method = "BH")) %>%
  dplyr::select(x, y, z, rho, p, n, message, q_x, q_y, q)
```

### save pcorr file

```{r}
write_csv(pcorr_sex_gmr, "~/Documents/gutmicrobiome_pharmacology/partial correlation gene and metab rectum cf sex adult updated 3-26-2026.csv")
```

## for alex

```{r}
a1 <- metadata_hxtx_rectum_yes %>%
  dplyr::select(External.ID, Participant.ID)

a2 <- metadata_metab_weeks %>%
  dplyr::select(External.ID, Participant.ID)

common_ids <- intersect(a1$Participant.ID, a2$Participant.ID)
a3 <- data.frame(Participant.ID = common_ids)

write_csv(a1, "~/Documents/gutmicrobiome_pharmacology/1-RNAseq_ParticipantIDs.csv")
write_csv(a2, "~/Documents/gutmicrobiome_pharmacology/2-metabolomics_ParticipantIDs.csv")
write_csv(a3, "~/Documents/gutmicrobiome_pharmacology/3-intersecting_ParticipantIDs.csv")
```
