---
title: "Figure 2 scatterplots of gene vs metab and metab vs regorafenib correlations"
output: html_document
date: "2025-07-04"
---

# Loading packages

```{r}
library(dplyr)
library(tidyverse)
library(scales)
library(svglite)
library(ggrastr)
library(ggrepel)
```

## theme across figures

```{r}
paper_theme <- theme_minimal() +
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
```

# load gene and metab files

```{r}
gmr_corr <- read.csv("~/Documents/gutmicrobiome_pharmacology/correlation gene and metab rectum adult updated 3-26-2026.csv")

gmr_corr1 <- gmr_corr |>
  dplyr::select(x, y, rho, p, q_x, q_y)
```

## cf diagnosis

```{r}
gmrd_pcorr <- read.csv("~/Documents/gutmicrobiome_pharmacology/partial correlation gene and metab rectum cf diagnosis adult updated 3-26-2026.csv")

gmrd_pcorr1 <- gmrd_pcorr |>
  dplyr::select(x, y, rho, p, q_x, q_y)
```

## cf sex

```{r}
gmrs_pcorr <- read.csv("~/Documents/gutmicrobiome_pharmacology/partial correlation gene and metab rectum cf sex adult updated 3-26-2026.csv")

gmrs_pcorr1 <- gmrs_pcorr |>
  dplyr::select(x, y, rho, p, q_x, q_y)
```

## adding in rhos to compare

```{r}
gmr_rho1 <- gmr_corr1 %>% 
  dplyr::rename(rho1 = rho)
gmrd_rho2 <- gmrd_pcorr1 %>% 
  dplyr::rename(rho2 = rho)
gmrs_rho3 <- gmrs_pcorr1 %>%
  dplyr::rename(rho3 = rho)

join_gmr_diag <- full_join(gmr_rho1, gmrd_rho2, by = c("x", "y"))

join_gmr_sex <- full_join(gmr_rho1, gmrs_rho3, by = c("x", "y"))
```

## isolating significant rhos (i.e. q < 0.05)

```{r}
fjoin_gmr_diag <- join_gmr_diag %>%
  dplyr::select(x, y, rho1, p.x, rho2, p.y, q_x.y, q_y.y) %>%  # keep q_x and q_y from partial
  dplyr::rename(q_x = q_x.y,    # FDR gene from partial spearman (diagnosis)
                q_y = q_y.y,    # FDR metab from partial spearman (diagnosis)
                p1 = p.x,       # p from raw spearman
                p2 = p.y)       # p from partial spearman

fjoin_gmr_sex <- join_gmr_sex %>%
  dplyr::select(x, y, rho1, p.x, rho3, p.y, q_x.y, q_y.y) %>%  # keep q_x and q_y from partial
  dplyr::rename(q_x = q_x.y,    # FDR gene from partial spearman (sex)
                q_y = q_y.y,    # FDR metab from partial spearman (sex)
                p1 = p.x,       # p from raw spearman
                p3 = p.y)       # p from partial spearman
```

## creating significance column 

```{r}
r1 <- gmr_rho1 %>%
  mutate(FDR_gene_sig = case_when(
    q_x < 0.05 ~ "Gene Significant",
    q_x >= 0.05 ~ "Gene Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  ))

d1 <- fjoin_gmr_diag %>%
  mutate(FDR_gene_sig = case_when(
    q_x < 0.05 ~ "Gene Significant",
    q_x >= 0.05 ~ "Gene Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  ))

s1 <- fjoin_gmr_sex %>%
  mutate(FDR_gene_sig = case_when(
    q_x < 0.05 ~ "Gene Significant",
    q_x >= 0.05 ~ "Gene Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  ))
```

### saving files

```{r}
write_csv(r1, "~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum FDR categorized 4-1-2026.csv")
write_csv(d1, "~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum diag FDR categorized 4-1-2026.csv")
write_csv(s1, "~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum sex FDR categorized 4-1-2026.csv")
```

### load files

```{r}
# load packages
library(tidyverse)
library(scales)
library(svglite)
library(ggrastr)

# paper theme
paper_theme <- theme_minimal() +
  theme(
    axis.title       = element_text(size = 20),
    axis.text        = element_text(size = 16),
    legend.title     = element_text(size = 20),
    legend.text      = element_text(size = 16),
    aspect.ratio     = 1
  )

# load files
s1 <- read.csv("~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum sex FDR categorized 4-1-2026.csv")
d1 <- read.csv("~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum diag FDR categorized 4-1-2026.csv")
r1 <- read.csv("~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum FDR categorized 4-1-2026.csv")
```

# Figure 2a: scatterplot sex

```{r}
p <- ggplot(s1, aes(x = rho1, y = rho3, color = FDR_metab_sig)) +
  rasterise(geom_point(size = 0.1), dpi = 300) +  # points are raster
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
  labs(
    x = "Spearman Correlation",
    y = "Partial Spearman Correlation controlling for Sex",
    color = "Significance"
  ) +
  scale_color_manual(values = c(
    "Metabolite Significant" = "#D55E00",
    "Metabolite Not Significant" = "#BBBBBB"
  )) +
  scale_x_continuous(labels = scales::label_number()) +
  paper_theme

ggsave("~/Documents/gutmicrobiome_pharmacology/figure2a_sex.svg",
       plot = p,
       device = svglite::svglite, 
       width = 8, height = 6,
       units = "in")
```

```{r}
p <- ggplot() +
  rasterise(
    geom_point(data = s1 %>% filter(FDR_metab_sig == "Metabolite Not Significant"),
               aes(x = rho1, y = rho3, color = "Metabolite Not Significant"),
               size = 0.1),
    dpi = 300
  ) +
  rasterise(
    geom_point(data = s1 %>% filter(FDR_metab_sig == "Metabolite Significant"),
               aes(x = rho1, y = rho3, color = "Metabolite Significant"),
               size = 0.3),
    dpi = 300
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
  labs(
    x     = "Spearman Correlation",
    y     = str_wrap("Partial Spearman Correlation controlling for Sex", width = 30),
    color = "Significance"
  ) +
  scale_color_manual(values = c(
    "Metabolite Significant"     = "#D55E00",
    "Metabolite Not Significant" = "#CCCCCC"
  )) +
  paper_theme

ggsave("~/Documents/gutmicrobiome_pharmacology/figure2a_sex.svg",
       plot = p,
       device = svglite::svglite,
       width = 8, height = 6,
       units = "in")

# save TIFF
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2a_sex.tiff",
       plot = p, device = "tiff",
       width = 8, height = 6, units = "in", dpi = 300)

# save PDF
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2a_sex.pdf",
       plot = p, device = "pdf",
       width = 8, height = 6, units = "in")
```

# Figure 2b: scatterplot diagnosis

```{r}
p2 <- ggplot(d1, aes(x = rho1, y = rho2, color = FDR_metab_sig)) +
  rasterise(geom_point(size = 0.1), dpi = 300) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
  labs(
    x = "Spearman Correlation",
    y = "Partial Spearman Correlation controlling for Disease Status",
    color = "Significance"
  ) +
  scale_color_manual(values = c(
    "Metabolite Significant" = "#D55E00",
    "Metabolite Not Significant" = "#BBBBBB"
  )) +
  scale_x_continuous(labels = scales::label_number()) +
  paper_theme

ggsave("~/Documents/gutmicrobiome_pharmacology/figure2b_diagnosis.svg",
       plot = p2,
       device = svglite::svglite, 
       width = 8, height = 6,
       units = "in")
```

```{r}
p2 <- ggplot() +
  rasterise(
    geom_point(data = d1 %>% filter(FDR_metab_sig == "Metabolite Not Significant"),
               aes(x = rho1, y = rho2, color = "Metabolite Not Significant"),
               size = 0.1),
    dpi = 300
  ) +
  rasterise(
    geom_point(data = d1 %>% filter(FDR_metab_sig == "Metabolite Significant"),
               aes(x = rho1, y = rho2, color = "Metabolite Significant"),
               size = 0.3),
    dpi = 300
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
  labs(
    x     = "Spearman Correlation",
    y     = str_wrap("Partial Spearman Correlation controlling for Disease Status", width = 30),
    color = "Significance"
  ) +
  scale_color_manual(values = c(
    "Metabolite Significant"     = "#D55E00",
    "Metabolite Not Significant" = "#CCCCCC"
  )) +
  paper_theme

ggsave("~/Documents/gutmicrobiome_pharmacology/figure2b_diagnosis.svg",
       plot = p2,
       device = svglite::svglite,
       width = 8, height = 6,
       units = "in")

# save TIFF
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2b_diagnosis.tiff",
       plot = p2, device = "tiff",
       width = 8, height = 6, units = "in", dpi = 300)

# save PDF
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2b_diagnosis.pdf",
       plot = p2, device = "pdf",
       width = 8, height = 6, units = "in")
```

# loading drug and metab files

```{r}
d <- read.csv("~/Documents/gutmicrobiome_pharmacology/correlation drug and metab rectum cf diagnosis adult updated 3-26-2026.csv")
s <- read.csv("~/Documents/gutmicrobiome_pharmacology/correlation drug and metab rectum cf sex adult updated 3-26-2026.csv")
r <- read.csv("~/Documents/gutmicrobiome_pharmacology/correlation drug and metab rectum full adult updated 3-26-2026.csv")
```

# isolating for regorafenib

```{r}
rr <- r %>%
  filter(x == "regorafenib")

rd <- d %>%
  filter(x == "regorafenib")

rs <- s %>%
  filter(x == "regorafenib")
```

## adding in rhos to compare

```{r}
rr_rho1 <- rr %>% 
  dplyr::rename(rho1 = rho)
rd_rho2 <- rd %>% 
  dplyr::rename(rho2 = rho)
rs_rho3 <- rs %>%
  dplyr::rename(rho3 = rho)

join_rd <- full_join(rr_rho1, rd_rho2, by = c("x", "y"))

join_rs <- full_join(rr_rho1, rs_rho3, by = c("x", "y"))

join_rd1 <- join_rd %>%
  dplyr::select(x, y, rho1, p.x, rho2, p.y, q_x.y, q_y.y) %>%
  dplyr::rename(q_x = `q_x.y`,
                q_y = `q_y.y`)
join_rs1 <- join_rs %>%
  dplyr::select(x, y, rho1, p.x, rho3, p.y, q_x.y, q_y.y) %>%
  dplyr::rename(q_x = `q_x.y`,
                q_y = `q_y.y`)

rr1 <- rr %>%
  dplyr::select(x, y, rho, p, q_x, q_y)
```

# adding significance categories

```{r}
rr_sig <- rr1 %>%
  mutate(FDR_drug_sig = case_when(
    q_x < 0.05 ~ "Drug Significant",
    q_x >= 0.05 ~ "Drug Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  )) %>%
  mutate(combined_significance = paste(FDR_drug_sig, FDR_metab_sig, sep = " & "))

rd_sig <- join_rd1 %>%
  mutate(FDR_drug_sig = case_when(
    q_x < 0.05 ~ "Drug Significant",
    q_x >= 0.05 ~ "Drug Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  )) %>%
  mutate(combined_significance = paste(FDR_drug_sig, FDR_metab_sig, sep = " & "))

rs_sig <- join_rs1 %>%
  mutate(FDR_drug_sig = case_when(
    q_x < 0.05 ~ "Drug Significant",
    q_x >= 0.05 ~ "Drug Not Significant"
  )) %>%
  mutate(FDR_metab_sig = case_when(
    q_y < 0.05 ~ "Metabolite Significant",
    q_y >= 0.05 ~ "Metabolite Not Significant"
  )) %>%
  mutate(combined_significance = paste(FDR_drug_sig, FDR_metab_sig, sep = " & "))
```

### saving files

```{r}
write_csv(rr_sig, "~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum FDR significance categorized 4-1-2026.csv")

write_csv(rd_sig, "~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum diag FDR significance categorized 4-1-2026.csv")

write_csv(rs_sig, "~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum sex FDR significance categorized 4-1-2026.csv")
```

# set up for labeling

```{r}
top_labels_d <- rd_sig %>%
  filter(combined_significance == "Drug Significant & Metabolite Significant" & abs(rho1) > 0.16 & abs(rho2) > 0.16) %>%
  arrange(rho1, rho2) %>%
  slice_head(n = 15)

top_labels_s <- rs_sig %>%
  filter(combined_significance == "Drug Significant & Metabolite Significant" & abs(rho1) > 0.16 & abs(rho3) > 0.16) %>%
  arrange(rho1, rho3) %>%
  slice_head(n = 15)

# Split y column into metabolite name and method/ID
top_labels_s_clean <- top_labels_s %>%
  mutate(
    metabolite = sub("_.*", "", y),          # everything before first underscore
    method_id  = sub("^[^_]*_", "", y)       # everything after first underscore
  )

top_labels_d_clean <- top_labels_d %>%
  mutate(
    metabolite = sub("_.*", "", y),
    method_id  = sub("^[^_]*_", "", y)
  )

# Reference table
ref_table <- top_labels_s_clean %>%
  dplyr::select(y, metabolite, method_id) %>%
  mutate(source = "sex") %>%
  bind_rows(
    top_labels_d_clean %>%
      dplyr::select(y, metabolite, method_id) %>%
      mutate(source = "diagnosis")
  ) %>%
  distinct()

write_csv(ref_table, "~/Documents/gutmicrobiome_pharmacology/metabolite_label_reference.csv")
```

# Figure 2c: scatterplot sex

```{r}
xr <- range(rs_sig$rho1, na.rm = TRUE)
yr <- range(rs_sig$rho3, na.rm = TRUE)
pad <- 0.02
xlims <- c(xr[1] - pad, xr[2] + pad)
ylims <- c(yr[1] - pad, yr[2] + pad)

segment.size = 0.4

ggplot(rs_sig, aes(x = rho1, y = rho3, color = combined_significance)) +
  geom_point(size = 1) +
  geom_label_repel(
    data = top_labels_s_clean,
    aes(label = metabolite),
    color = "black",
    fill = "white",
    size = 3.5,
    box.padding = 0.2,
    point.padding = 0.2,
    segment.color = "gray60",
    segment.linewidth = 0.4,
    label.size = 0.25,
    label.r = unit(0.1, "lines"),
    force = 50,
    max.overlaps = Inf,
    min.segment.length = 0,
    seed = 42              # reproducible label placement
  ) +
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +  # fixed
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +  # fixed
  coord_cartesian(xlim = xlims, ylim = ylims, expand = FALSE) +
  scale_color_manual(values = c(
    "Drug Significant & Metabolite Significant" = "#D55E00",
    "Drug Not Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Not Significant & Metabolite Significant" = "#BBBBBB"
  )) +
  labs(
    x = "Spearman Correlation",
    y = "Partial Spearman Correlation controlling for Sex",
    color = "Significance"
  ) +
  paper_theme +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white"),
    axis.ticks = element_line(color = "black")
  )
```

# Figure 2d: scatterplot diagnosis

```{r}
ggplot(rd_sig, aes(x = rho1, y = rho2, color = combined_significance)) +
  geom_point(size = 1) +
  geom_label_repel(
    data = top_labels_d_clean,        # updated to clean version
    aes(label = metabolite),          # just metabolite name
    color = "black",
    fill = "white",
    size = 3.5,                       # updated
    box.padding = 0.2,                # updated
    point.padding = 0.2,              # updated
    segment.color = "gray60",
    segment.linewidth = 0.4,          # fixed deprecation
    label.size = 0.25,
    label.r = unit(0.1, "lines"),
    force = 50,                       # updated
    max.overlaps = Inf,
    min.segment.length = 0,           # added
    seed = 42                         # added
  ) +
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +  # updated
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +  # updated
  coord_cartesian(xlim = xlims, ylim = ylims, expand = FALSE) +
  scale_color_manual(values = c(
    "Drug Significant & Metabolite Significant" = "#D55E00",
    "Drug Not Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Not Significant & Metabolite Significant" = "#BBBBBB"
  )) +
  labs(
    x = "Spearman Correlation",
    y = "Partial Spearman Correlation controlling for Disease Status",  # updated
    color = "Significance"
  ) +
  paper_theme +                       # updated
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white"),
    axis.ticks = element_line(color = "black")
  )
```

```{r}
# load files
rs_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum sex FDR significance categorized 4-1-2026.csv")
rd_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum diag FDR significance categorized 4-1-2026.csv")
rr_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum FDR significance categorized 4-1-2026.csv")
```

```{r}
# label setup
top_labels_d <- rd_sig %>%
  filter(combined_significance == "Drug Significant & Metabolite Significant" & 
           abs(rho1) > 0.16 & abs(rho2) > 0.16) %>%
  arrange(rho1, rho2) %>%
  slice_head(n = 15)

top_labels_s <- rs_sig %>%
  filter(combined_significance == "Drug Significant & Metabolite Significant" & 
           abs(rho1) > 0.16 & abs(rho3) > 0.16) %>%
  arrange(rho1, rho3) %>%
  slice_head(n = 15)

top_labels_s_clean <- top_labels_s %>%
  mutate(
    metabolite = sub("_.*", "", y),
    method_id  = sub("^[^_]*_", "", y)
  )

top_labels_d_clean <- top_labels_d %>%
  mutate(
    metabolite = sub("_.*", "", y),
    method_id  = sub("^[^_]*_", "", y)
  )
```

```{r}
# Figure 2c: scatterplot sex
xr_s <- range(rs_sig$rho1, na.rm = TRUE)
yr_s <- range(rs_sig$rho3, na.rm = TRUE)
pad  <- 0.02
xlims_s <- c(xr_s[1] - pad, xr_s[2] + pad)
ylims_s <- c(yr_s[1] - pad, yr_s[2] + pad)

p3 <- ggplot(rs_sig, aes(x = rho1, y = rho3, color = combined_significance)) +
  geom_point(size = 1) +
  geom_label_repel(
    data = top_labels_s_clean,
    aes(label = metabolite),
    color = "black", fill = "white",
    size = 4.5, box.padding = 0.2, point.padding = 0.2,
    segment.color = "gray60", segment.linewidth = 0.4,
    label.size = 0.25, label.r = unit(0.1, "lines"),
    force = 50, max.overlaps = Inf,
    min.segment.length = 0, seed = 42
  ) +
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +
  coord_cartesian(xlim = xlims_s, ylim = ylims_s, expand = FALSE) +
  scale_color_manual(values = c(
    "Drug Significant & Metabolite Significant"         = "#D55E00",
    "Drug Not Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Significant & Metabolite Not Significant"     = "#BBBBBB",
    "Drug Not Significant & Metabolite Significant"     = "#BBBBBB"
  )) +
  labs(
    x     = "Spearman Correlation",
    y     = str_wrap("Partial Spearman Correlation controlling for Sex", width = 30),
    color = "Significance"
  ) +
  paper_theme +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white"),
    axis.ticks       = element_line(color = "black")
  )

# save
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2c_sex.svg",
       plot = p3, device = svglite::svglite,
       width = 12, height = 8, units = "in")
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2c_sex.tiff",
       plot = p3, device = "tiff",
       width = 8, height = 6, units = "in", dpi = 300)
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2c_sex.pdf",
       plot = p3, device = "pdf",
       width = 8, height = 6, units = "in")
```

```{r}
# Figure 2d: scatterplot diagnosis — separate axis limits from rd_sig
xr_d <- range(rd_sig$rho1, na.rm = TRUE)
yr_d <- range(rd_sig$rho2, na.rm = TRUE)
xlims_d <- c(xr_d[1] - pad, xr_d[2] + pad)
ylims_d <- c(yr_d[1] - pad, yr_d[2] + pad)

p4 <- ggplot(rd_sig, aes(x = rho1, y = rho2, color = combined_significance)) +
  geom_point(size = 1) +
  geom_label_repel(
    data = top_labels_d_clean,
    aes(label = metabolite),
    color = "black", fill = "white",
    size = 4.5, box.padding = 0.2, point.padding = 0.2,
    segment.color = "gray60", segment.linewidth = 0.4,
    label.size = 0.25, label.r = unit(0.1, "lines"),
    force = 50, max.overlaps = Inf,
    min.segment.length = 0, seed = 42
  ) +
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +
  coord_cartesian(xlim = xlims_d, ylim = ylims_d, expand = FALSE) +  # diagnosis limits
  scale_color_manual(values = c(
    "Drug Significant & Metabolite Significant"         = "#D55E00",
    "Drug Not Significant & Metabolite Not Significant" = "#BBBBBB",
    "Drug Significant & Metabolite Not Significant"     = "#BBBBBB",
    "Drug Not Significant & Metabolite Significant"     = "#BBBBBB"
  )) +
  labs(
    x     = "Spearman Correlation",
    y     = str_wrap("Partial Spearman Correlation controlling for Disease Status", width = 30),
    color = "Significance"
  ) +
  paper_theme +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white"),
    axis.ticks       = element_line(color = "black")
  )

# save
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2d_diagnosis.svg",
       plot = p4, device = svglite::svglite,
       width = 12, height = 8, units = "in")
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2d_diagnosis.tiff",
       plot = p4, device = "tiff",
       width = 8, height = 6, units = "in", dpi = 300)
ggsave("~/Documents/gutmicrobiome_pharmacology/figure2d_diagnosis.pdf",
       plot = p4, device = "pdf",
       width = 8, height = 6, units = "in")
```
