---
title: "7 - Fig 4 heatmaps of top metabolites and pathways"
output: html_document
date: "2025-09-26"
---

# load packages 

```{r}
library(arrow)
library(tidyverse)
library(pheatmap)
library(KEGGREST)
library(clusterProfiler)
library(org.Hs.eg.db)
library(fgsea)
```

# load files

```{r}
x <- arrow::read_parquet("lincs_drug_gene_data_mean")
xmeta <- arrow::read_parquet("lincs_drug_meta")

gmr_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum FDR categorized 4-1-2026.csv")

rr_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/regorafenib vs metab correlations adult rectum FDR significance categorized 4-1-2026.csv")

mimics_metabs    <- read.csv("~/Documents/gutmicrobiome_pharmacology/mimics_metabs.csv")
opposites_metabs <- read.csv("~/Documents/gutmicrobiome_pharmacology/opposites_metabs.csv")

kegg_gene_sets <- readRDS("~/Documents/gutmicrobiome_pharmacology/kegg_gene_sets.rds")

overlap_mimics <- mimics_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>%
  unique() %>%
  na.omit()

overlap_opposites <- opposites_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>%
  unique() %>%
  na.omit()

# shared function
format_metab_name <- function(x) {
  ifelse(is.na(x), NA,
    paste0(sub("_.*", "", x), " (", sub("^[^_]*_", "", x), ")")
  )
}

# clean pathway names function
clean_pathway_names <- function(mat) {
  rownames(mat) <- gsub(" - Homo sapiens \\(human\\)", "", rownames(mat))
  mat
}
```

# filter for regorafenib

```{r}
regorafenib <- x %>%
  dplyr::select(gene, regorafenib) %>%
  dplyr::rename(x = gene)
```

## regorafenib gene perturbation

```{r}
write_csv(regorafenib, "~/Documents/gutmicrobiome_pharmacology/regorafenib gene pertubation.csv")
```

# FDR scores for regorafenib

```{r}
reg_z <- regorafenib %>%
  mutate(
    z = as.numeric(scale(regorafenib)),
    pval = 2 * pnorm(-abs(z)),
    FDR = p.adjust(pval, method = "fdr")
  )
```

## save FDR scores

```{r}
write_csv(reg_z, "~/Documents/gutmicrobiome_pharmacology/regorafenib gene pertubation with FDR.csv")
```

# filtering top mimics and opposites from raw spearman gene-metab

```{r}
overlap_mimics <- mimics_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>%
  unique() %>%
  na.omit()

overlap_opposites <- opposites_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>%
  unique() %>%
  na.omit()

top_mimics <- rr_sig %>%
  filter(y %in% overlap_mimics) %>%
  dplyr::select(y, rho, q_x, q_y, FDR_drug_sig, FDR_metab_sig) %>%
  mutate(Metabolite = format_metab_name(y)) %>%
  dplyr::select(Metabolite, rho, q_x, q_y, FDR_drug_sig, FDR_metab_sig) %>%
  dplyr::rename(
    "Unadjusted rho"          = rho,
    "FDR (drug)"              = q_x,
    "FDR (metabolite)"        = q_y,
    "Drug significance"       = FDR_drug_sig,
    "Metabolite significance" = FDR_metab_sig
  )

top_opposites <- rr_sig %>%
  filter(y %in% overlap_opposites) %>%
  dplyr::select(y, rho, q_x, q_y, FDR_drug_sig, FDR_metab_sig) %>%
  mutate(Metabolite = format_metab_name(y)) %>%
  dplyr::select(Metabolite, rho, q_x, q_y, FDR_drug_sig, FDR_metab_sig) %>%
  dplyr::rename(
    "Unadjusted rho"          = rho,
    "FDR (drug)"              = q_x,
    "FDR (metabolite)"        = q_y,
    "Drug significance"       = FDR_drug_sig,
    "Metabolite significance" = FDR_metab_sig
  )
```

## save top metabolites

```{r}
write_csv(top_mimics, "~/Documents/gutmicrobiome_pharmacology/top_mimics_regorafenib_metabolites_4-2-2026.csv")
write_csv(top_opposites, "~/Documents/gutmicrobiome_pharmacology/top_opposites_regorafenib_metabolites_4-2-2026.csv")
```

# function over multiple metabolites

```{r}
# fgsea function - WITH formatted metabolite names
run_fgsea_for_metabolite <- function(met_name, gmr_sig, reg_z, pathways) {
  df <- gmr_sig %>%
    filter(y == met_name) %>%
    left_join(reg_z, by = "x") %>%
    na.omit() %>%
    mutate(combined_score = sign(rho1) * -log10(pval))
  ranks <- setNames(df$combined_score, df$x) %>% sort(decreasing = TRUE)
  fg <- fgsea(pathways = pathways, stats = ranks, minSize = 15, maxSize = 500, nPermSimple = 10000)
  if (nrow(fg) == 0) return(NULL)
  as_tibble(fg) %>%
    mutate(metabolite = format_metab_name(met_name))  # formatted name here
}
```

# run fgsea

```{r}
fgsea_mimics <- map_dfr(overlap_mimics, run_fgsea_for_metabolite, gmr_sig, reg_z, kegg_gene_sets)
fgsea_opposites <- map_dfr(overlap_opposites, run_fgsea_for_metabolite, gmr_sig, reg_z, kegg_gene_sets)

write_csv(fgsea_mimics, "~/Documents/gutmicrobiome_pharmacology/fgsea_top_mimics_regorafenib_metabolites.csv")
write_csv(fgsea_opposites, "~/Documents/gutmicrobiome_pharmacology/fgsea_top_opposites_regorafenib_metabolites.csv")
```

# setting up heatmap

## filter for padj <0.05

```{r}
# For mimics
fgsea_mat_mimics <- fgsea_mimics %>%
  filter(padj < 0.05) %>%
  dplyr::select(pathway, metabolite, NES) %>%
  pivot_wider(names_from = metabolite, values_from = NES, values_fill = 0) %>%
  column_to_rownames("pathway") %>%
  clean_pathway_names()

# For opposites
fgsea_mat_opposites <- fgsea_opposites %>%
  filter(padj < 0.05) %>%
  dplyr::select(pathway, metabolite, NES) %>%
  pivot_wider(names_from = metabolite, values_from = NES, values_fill = 0) %>%
  column_to_rownames("pathway") %>%
  clean_pathway_names()

# colorblind safe palette
my_palette <- colorRampPalette(c("#2166AC", "white", "#D55E00"))(100)

# convert structural zeros to NA before plotting
fgsea_mat_mimics_plot <- as.matrix(fgsea_mat_mimics)
fgsea_mat_mimics_plot[fgsea_mat_mimics_plot == 0] <- NA

fgsea_mat_opposites_plot <- as.matrix(fgsea_mat_opposites)
fgsea_mat_opposites_plot[fgsea_mat_opposites_plot == 0] <- NA

# Figure 4 heatmap - mimics
pheatmap(fgsea_mat_mimics_plot,
         cluster_rows  = row_clust_m,
         cluster_cols  = col_clust_m,
         scale         = "none",        # no scaling — use raw NES values
         color         = my_palette,
         na_col        = "white",
         fontsize      = 12,
         fontsize_row  = 10,
         fontsize_col  = 11,
         main          = "")

# Figure 5 heatmap - opposites
pheatmap(fgsea_mat_opposites_plot,
         cluster_rows  = row_clust_o,
         cluster_cols  = col_clust_o,
         scale         = "none",        # no scaling — use raw NES values
         color         = my_palette,
         na_col        = "white",
         fontsize      = 12,
         fontsize_row  = 10,
         fontsize_col  = 11,
         main          = "")
```

## save matrices

```{r}
write.csv(fgsea_mat_mimics, "~/Documents/gutmicrobiome_pharmacology/fgsea_matrices_top_mimics_regorafenib_metabolites-4-8-2026.csv")
write.csv(fgsea_mat_opposites, "~/Documents/gutmicrobiome_pharmacology/fgsea_matrices_top_opposites_regorafenib_metabolites-4-8-2026.csv")
```

# get KEGG pathway IDs

```{r}
kegg_pathway_ids <- keggList("pathway", "hsa")
```

# get gene sets

```{r}
kegg_gene_sets <- lapply(seq_along(kegg_pathway_ids), function(i) {
  pid <- names(kegg_pathway_ids)[i]
  tryCatch({
    result <- keggGet(pid)[[1]]
    genes <- result$GENE
    if (is.null(genes)) return(NULL)
    # odd indices are entrez IDs, even are descriptions
    entrez_ids <- genes[seq(1, length(genes), by = 2)]
    # convert entrez to gene symbols to match your x column in gmr_sig
    symbols <- mapIds(org.Hs.eg.db, 
                      keys = entrez_ids, 
                      column = "SYMBOL", 
                      keytype = "ENTREZID",
                      multiVals = "first")
    symbols <- symbols[!is.na(symbols)]
    if (length(symbols) == 0) return(NULL)
    unname(symbols)
  }, error = function(e) NULL)
})
names(kegg_gene_sets) <- kegg_pathway_ids
kegg_gene_sets <- Filter(Negate(is.null), kegg_gene_sets)
```

```{r}
saveRDS(kegg_gene_sets, "~/Documents/gutmicrobiome_pharmacology/kegg_gene_sets.rds")
```

```{r}
# function over multiple metabolites
run_fgsea_for_metabolite <- function(met_name, gmr_sig, reg_z, pathways) {
  df <- gmr_sig %>%
    filter(y == met_name) %>%
    left_join(reg_z, by = "x") %>%
    na.omit() %>%
    mutate(combined_score = sign(rho1) * -log10(pval))
  ranks <- setNames(df$combined_score, df$x) %>% sort(decreasing = TRUE)
  fg <- fgsea(pathways = pathways, stats = ranks, minSize = 15, maxSize = 500, nPermSimple = 10000)
  if (nrow(fg) == 0) return(NULL)
  as_tibble(fg) %>%
    mutate(metabolite = met_name,
           metabolite = format_metab_name(met_name))  # format name here
}

# run fgsea
fgsea_mimics <- map_dfr(overlap_mimics, run_fgsea_for_metabolite, gmr_sig, reg_z, kegg_gene_sets)
fgsea_opposites <- map_dfr(overlap_opposites, run_fgsea_for_metabolite, gmr_sig, reg_z, kegg_gene_sets)
```

# save fgsea files

```{r}
write_csv(fgsea_mimics,    "~/Documents/gutmicrobiome_pharmacology/fgsea_top_mimics_regorafenib_metabolites-4-2-2026.csv")
write_csv(fgsea_opposites, "~/Documents/gutmicrobiome_pharmacology/fgsea_top_opposites_regorafenib_metabolites-4-2-2026.csv")
```
