---
title: "Figure 6 significant genes"
output: html_document
date: "2025-09-09"
---

# load packages and files

```{r}
library(tidyverse)
library(survival)
library(survminer)

gmr_sig <- read.csv("~/Documents/gutmicrobiome_pharmacology/gene vs metab correlations adult rectum FDR categorized 4-1-2026.csv")

reg_z <- read.csv("~/Documents/gutmicrobiome_pharmacology/regorafenib gene pertubation with FDR.csv")

top_mimics <- read.csv("~/Documents/gutmicrobiome_pharmacology/top_mimics_regorafenib_metabolites_4-2-2026.csv")
top_opposites <- read.csv("~/Documents/gutmicrobiome_pharmacology/top_opposites_regorafenib_metabolites_4-2-2026.csv")
```

# filter for genes with significance + consistent direction

```{r}
# function to process one metabolite
process_metabolite <- function(met_name, gmr_sig, reg_z) {
  gmr_sig %>%
    filter(y == met_name) %>%
    left_join(reg_z, by = "x") %>%
    na.omit() %>%
    dplyr::select(x, y, rho1, p, q_x, q_y, pval, FDR) %>% 
    filter(p < 0.05, FDR < 0.05)
}

mimics_metabs    <- read.csv("~/Documents/gutmicrobiome_pharmacology/mimics_metabs.csv")
opposites_metabs <- read.csv("~/Documents/gutmicrobiome_pharmacology/opposites_metabs.csv")

overlap_mimics <- mimics_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>% unique() %>% na.omit()

overlap_opposites <- opposites_metabs %>%
  filter(category == "Common to all three analyses") %>%
  pull(y) %>% unique() %>% na.omit()

mimics_results <- map(overlap_mimics, ~process_metabolite(.x, gmr_sig, reg_z)) %>%
  bind_rows() %>%
  mutate(group = "mimic")

opposites_results <- map(overlap_opposites, ~process_metabolite(.x, gmr_sig, reg_z)) %>%
  bind_rows() %>%
  mutate(group = "opposite")
```

# data from cBioPortal - Colorectal Adenocarcinoma (TCGA, PanCancer Atlas)

```{r}
#clinical data
clin <- read.delim("~/Documents/gutmicrobiome_pharmacology/coadread_tcga_pan_can_atlas_2018/data_clinical_patient.txt", 
                   comment.char = "#")

#mrna expression z-scores
expr <- read.delim(
  "~/Documents/gutmicrobiome_pharmacology/coadread_tcga_pan_can_atlas_2018/data_mrna_seq_v2_rsem_zscores_ref_all_samples.txt",
  check.names = FALSE
)
```

### clean expr

```{r}
expr$Hugo_Symbol <- ifelse(expr$Hugo_Symbol == "" | is.na(expr$Hugo_Symbol), 
                           expr$Entrez_Gene_Id, 
                           expr$Hugo_Symbol)

expr_median <- expr %>%
  rowwise() %>%
  mutate(
    row_median = median(c_across(!any_of(c("Hugo_Symbol", "Entrez_Gene_Id"))), na.rm = TRUE)
  ) %>%
  ungroup()

# Keep the row with the highest median per Hugo_Symbol
expr_unique <- expr_median %>%
  group_by(Hugo_Symbol) %>%
  slice_max(row_median, n = 1) %>%
  ungroup() %>%
  dplyr::select(-row_median)  # remove helper column

expr_unique_matrix <- expr_unique %>%
  column_to_rownames("Hugo_Symbol") %>%
  dplyr::select(-Entrez_Gene_Id)

# Trim expression sample IDs to match clinical PATIENT_IDs
colnames(expr_unique_matrix) <- sub("-01$", "", colnames(expr_unique_matrix))
```

## standardize survival variables

```{r}
clin$OS_months <- as.numeric(clin$OS_MONTHS)
clin$OS_status <- as.numeric(sub(":.*", "", clin$OS_STATUS))
```

## confirm sample IDs match

```{r}
common_ids <- intersect(colnames(expr_unique_matrix), clin$PATIENT_ID)

clin2 <- clin[match(common_ids, clin$PATIENT_ID), ]
expr2 <- expr_unique_matrix[, common_ids]
```

# setting up KM plots

```{r}
shared_genes <- intersect(mimics_results$x, opposites_results$x)

cat("Genes excluded due to mixed mimic/opposite associations:", length(shared_genes), "\n")
print(shared_genes)

mimics_results_clean    <- mimics_results %>% filter(!x %in% shared_genes)
opposites_results_clean <- opposites_results %>% filter(!x %in% shared_genes)
```

## split gene lists b/w mimics and opposites

```{r}
mimic_genes    <- unique(mimics_results_clean$x)    # 15 genes
opposite_genes <- unique(opposites_results_clean$x) # 7 genes

cat("Mimic genes after exclusion:", length(mimic_genes), "\n")
cat("Opposite genes after exclusion:", length(opposite_genes), "\n")
```

## KM function to return p values and hazard ratios per gene

```{r}
analyze_gene <- function(gene, expr_data, clin_data, outdir = "KM_plots") {
  if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
  
  # get expression values for this gene
  if (!(gene %in% rownames(expr_data))) return(NULL)
  gene_expr <- as.numeric(expr_data[gene, ])
  
  # align clinical data
  df <- clin_data
  df$expr <- gene_expr
  
  # split by median
  df$group <- factor(ifelse(df$expr > median(df$expr, na.rm = TRUE), "High", "Low"), 
                     levels = c("Low", "High"))
  
  # Kaplan–Meier fit
  fit <- survfit(Surv(OS_months, OS_status) ~ group, data = df)
  
  # Cox model
  cox_fit <- coxph(Surv(OS_months, OS_status) ~ group, data = df)
  cox_sum <- summary(cox_fit)
  
  hr <- cox_sum$coefficients[,"exp(coef)"]
  pval <- cox_sum$coefficients[,"Pr(>|z|)"]
  
  # KM plot with HR and p-value
  g <- ggsurvplot(
    fit, data = df,
    risk.table    = TRUE,
    pval          = TRUE,
    conf.int      = FALSE,
    title         = "",
    legend.title  = "Expression",
    legend.labs   = c("Low", "High"),
    font.main     = 14,
    font.x        = 14,
    font.y        = 14,
    font.tickslab = 12,
    font.legend   = 12,
    risk.table.fontsize = 4,
    xlab     = "Time (months)",    # <-- add this
    ylab     = "Overall survival", # <-- optional but cleaner than default
    subtitle = paste0(gene, "  |  HR = ", round(hr, 2),
                      "  |  p = ", signif(pval, 3))
  )
  
  # save plot
  ggsave(filename = file.path(outdir, paste0(gene, "_KM.png")),
         plot = g$plot, width = 6, height = 5, dpi = 300)
  
  return(data.frame(Gene = gene, HR = hr, Pval = pval))
}
```

## loop mimics

```{r}
mimics_summary <- do.call(rbind, lapply(mimic_genes, function(g) {
  analyze_gene(g, expr2, clin2, outdir = "~/Documents/gutmicrobiome_pharmacology/KM_plots_mimics")
})) %>% arrange(Pval)
```

## loop opposites

```{r}
opposites_summary <- do.call(rbind, lapply(opposite_genes, function(g) {
  analyze_gene(g, expr2, clin2, outdir = "~/Documents/gutmicrobiome_pharmacology/KM_plots_opposites")
})) %>% arrange(Pval)
```

# save summary KM csvs

```{r}
write.csv(mimics_summary, "~/Documents/gutmicrobiome_pharmacology/mimics_KM_summary_4-3-2026.csv", row.names = FALSE)
write.csv(opposites_summary, "~/Documents/gutmicrobiome_pharmacology/opposites_KM_summary_4-3-2026.csv", row.names = FALSE)
```

# unique metabolites

```{r}
mimic_unique <- unique(mimics_results_clean$y)       
opposite_unique <- unique(opposites_results_clean$y) 

length(mimic_unique)    # 14
length(opposite_unique) # 5
```

# just HIST1H2AC

```{r}
# load packages
library(tidyverse)
library(survival)
library(survminer)
library(svglite)

# generate HIST1H2AC plot specifically
df_hist <- clin2
df_hist$expr  <- as.numeric(expr2["HIST1H2AC", ])
df_hist$group <- factor(ifelse(df_hist$expr > median(df_hist$expr, na.rm = TRUE), 
                                "High", "Low"),
                         levels = c("Low", "High"))

fit_hist     <- survfit(Surv(OS_months, OS_status) ~ group, data = df_hist)
cox_hist     <- coxph(Surv(OS_months, OS_status) ~ group, data = df_hist)
cox_sum_hist <- summary(cox_hist)
hr_hist      <- cox_sum_hist$coefficients[, "exp(coef)"]
pval_hist    <- cox_sum_hist$coefficients[, "Pr(>|z|)"]

g_hist <- ggsurvplot(
  fit_hist, data = df_hist,
  risk.table          = TRUE,
  pval                = TRUE,
  conf.int            = FALSE,
  title               = "",
  legend.title        = "Expression",
  legend.labs         = c("Low", "High"),
  font.main           = 14,
  font.x              = 14,
  font.y              = 14,
  font.tickslab       = 12,
  font.legend         = 12,
  risk.table.fontsize = 4,
  xlab                = "Time (months)",
  ylab                = "Overall survival",
  subtitle            = paste0("HIST1H2AC  |  HR = ", round(hr_hist, 2),
                               "  |  p = ", signif(pval_hist, 3))
)

# save TIFF
tiff("~/Documents/gutmicrobiome_pharmacology/figure6_HIST1H2AC_KM.tiff",
     width = 6, height = 7, units = "in", res = 300)
print(g_hist)
dev.off()

# save PDF
pdf("~/Documents/gutmicrobiome_pharmacology/figure6_HIST1H2AC_KM.pdf",
    width = 6, height = 7)
print(g_hist)
dev.off()

# save SVG
svglite::svglite("~/Documents/gutmicrobiome_pharmacology/figure6_HIST1H2AC_KM.svg",
                  width = 6, height = 7)
print(g_hist)
dev.off()
```
