---
title: "6 - Supp Fig 1 and 2"
output: html_document
date: "2025-10-01"
---

# load packages and files

```{r}
library(tidyverse)
library(ComplexUpset)
library(svglite)

fgsea_mimics <- read.csv("~/Documents/gutmicrobiome_pharmacology/fgsea_top_mimics_regorafenib_metabolites-4-2-2026.csv")
fgsea_opposites <- read.csv("~/Documents/gutmicrobiome_pharmacology/fgsea_top_opposites_regorafenib_metabolites-4-2-2026.csv")

format_metab_name <- function(x) {
  ifelse(is.na(x), NA,
    paste0(sub("_.*", "", x), " (", sub("^[^_]*_", "", x), ")")
  )
}

# paper theme
paper_theme <- theme_minimal() +
  theme(
    axis.title       = element_text(size = 16),
    axis.text        = element_text(size = 14),
    legend.title     = element_text(size = 16),
    legend.text      = element_text(size = 14)
  )
```

# setting up for plots

```{r}
# clean pathway names
clean_pathway <- function(x) gsub(" - Homo sapiens \\(human\\)", "", x)

# filter significant, add direction, clean pathway + metabolite names
fgsea_mimics_sig <- fgsea_mimics %>%
  filter(padj < 0.05) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway   = clean_pathway(pathway),
    #metabolite = format_metab_name(metabolite)
  )

fgsea_opposites_sig <- fgsea_opposites %>%
  filter(padj < 0.05) %>%
  mutate(
    direction  = ifelse(NES > 0, "Positive", "Negative"),
    pathway    = clean_pathway(pathway),
    #metabolite = format_metab_name(metabolite)
  )

# Create binary long table for upset plots
fgsea_mimics_bin <- fgsea_mimics_sig %>%
  dplyr::select(pathway, metabolite) %>%
  mutate(sig = 1) %>%
  pivot_wider(
    names_from = metabolite,
    values_from = sig,
    values_fill = 0
  )

fgsea_opposites_bin <- fgsea_opposites_sig %>%
  dplyr::select(pathway, metabolite) %>%
  mutate(sig = 1) %>%
  pivot_wider(
    names_from = metabolite,
    values_from = sig,
    values_fill = 0
  )
```

# Fig S1a - upset plot

```{r}
# mimics
ComplexUpset::upset(
  fgsea_mimics_bin,
  intersect = colnames(fgsea_mimics_bin)[-1],
  min_size  = 1,
  base_annotations = list(
    'Intersection size' = intersection_size(
      counts = TRUE,
      fill   = "steelblue",
      text   = list(size = 4)
    )
  )
) & theme(axis.text = element_text(size = 12),
          axis.title = element_text(size = 14))
```

# Fig S2a - upset plot

```{r}
# opposites
ComplexUpset::upset(
  fgsea_opposites_bin,
  intersect = colnames(fgsea_opposites_bin)[-1],
  min_size  = 1,
  base_annotations = list(
    'Intersection size' = intersection_size(
      counts = TRUE,
      fill   = "steelblue",
      text   = list(size = 4)
    ) + scale_y_continuous(breaks = scales::breaks_pretty())  # whole numbers
  ),
  themes = upset_modify_themes(list(
    'overall_sizes' = theme(axis.text = element_text(size = 12)),
    'intersections_matrix' = theme(axis.text = element_text(size = 12))
  ))
)
```

# supplementary tables — intersection pathway mapping (new code)

```{r}
mimics_metabolite_cols    <- colnames(fgsea_mimics_bin)[-1]
opposites_metabolite_cols <- colnames(fgsea_opposites_bin)[-1]

make_intersection_table <- function(bin_df, metabolite_cols, outfile) {
  bin_df %>%
    rowwise() %>%
    mutate(
      intersection = paste(which(c_across(all_of(metabolite_cols)) == 1), collapse = "-"),
      metabolites_in_intersection = paste(metabolite_cols[which(c_across(all_of(metabolite_cols)) == 1)], collapse = ", ")
    ) %>%
    ungroup() %>%
    group_by(intersection, metabolites_in_intersection) %>%
    summarise(
      num_pathways = n(),
      pathways     = paste(unique(pathway), collapse = "; "),
      .groups      = "drop"
    ) %>%
    mutate(
      num_metabolites_in_intersection = str_count(intersection, "-") + 1,
      overlap_type = ifelse(num_metabolites_in_intersection > 1, "Shared", "Single")
    ) %>%
    arrange(desc(num_pathways)) %>%
    write.csv(outfile, row.names = FALSE)
}

make_intersection_table(
  fgsea_mimics_bin, mimics_metabolite_cols,
  "~/Documents/gutmicrobiome_pharmacology/supp_fig1a_intersection_pathway_names.csv"
)

make_intersection_table(
  fgsea_opposites_bin, opposites_metabolite_cols,
  "~/Documents/gutmicrobiome_pharmacology/supp_fig2a_intersection_pathway_names.csv"
)
```

# separate stacked bar charts

## compute counts by direction for top bars

```{r}
compute_counts_upset_order <- function(bin_df, long_df) {
  metabolites <- bin_df %>% dplyr::select(where(is.numeric)) %>% colnames()
  bin_df <- bin_df %>%
    rowwise() %>%
    mutate(intersection = paste(which(c_across(all_of(metabolites)) == 1), collapse = "-")) %>%
    ungroup()
  direction_map <- long_df %>% dplyr::select(pathway, direction) %>% distinct()
  counts <- bin_df %>%
    left_join(direction_map, by = "pathway") %>%
    group_by(intersection, direction) %>%
    summarise(count = n_distinct(pathway), .groups = "drop")
  intersection_order <- counts %>%
    group_by(intersection) %>%
    summarise(total = sum(count)) %>%
    arrange(desc(total)) %>%
    pull(intersection)
  counts %>% mutate(intersection = factor(intersection, levels = intersection_order))
}

mimics_counts    <- compute_counts_upset_order(fgsea_mimics_bin,    fgsea_mimics_sig)
opposites_counts <- compute_counts_upset_order(fgsea_opposites_bin, fgsea_opposites_sig)
```

# Supp Fig 1b & 2b: plot stacked bar chart for top bars

## mimics

```{r}
ggplot(mimics_counts, aes(x = intersection, y = count, fill = direction)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Positive" = "#D82632", "Negative" = "#2166AC")) +
  labs(
    x    = "Intersection (metabolites numbered as in upset plot)",
    y    = "Number of pathways",
    fill = "NES direction"
  ) +
  paper_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 11))
```

## opposites

```{r}
ggplot(opposites_counts, aes(x = intersection, y = count, fill = direction)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Positive" = "#D82632", "Negative" = "#2166AC")) +
  scale_y_continuous(breaks = scales::breaks_pretty()) +   # whole numbers only
  labs(
    x    = "Intersection (metabolites numbered as in upset plot)",
    y    = "Number of pathways",
    fill = "NES direction"
  ) +
  paper_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 11))
```

## legend csv

```{r}
make_legend_csv <- function(counts_df, bin_df, metabolite_cols, outfile) {
  intersection_map <- bin_df %>%
    rowwise() %>%
    mutate(
      intersection = paste(which(c_across(all_of(metabolite_cols)) == 1), collapse = "-"),
      metabolites_in_intersection = paste(metabolite_cols[which(c_across(all_of(metabolite_cols)) == 1)], collapse = ", ")
    ) %>%
    ungroup() %>%
    dplyr::select(intersection, metabolites_in_intersection) %>%
    distinct()

  counts_df %>%
    dplyr::select(intersection, direction, count) %>%
    distinct() %>%
    left_join(intersection_map, by = "intersection") %>%
    arrange(intersection, direction) %>%
    write.csv(outfile, row.names = FALSE)
}

make_legend_csv(mimics_counts,    fgsea_mimics_bin,    mimics_metabolite_cols,
                "~/Documents/gutmicrobiome_pharmacology/supp_fig1b_mimics_top_bar_legend.csv")
make_legend_csv(opposites_counts, fgsea_opposites_bin, opposites_metabolite_cols,
                "~/Documents/gutmicrobiome_pharmacology/supp_fig2b_opposites_top_bar_legend.csv")
```

# Supp Fig 1c & 2c: plot stacked bar chart for side bars (per metabolite)

```{r}
make_side_bar <- function(sig_df) {
  sig_df %>%
    group_by(metabolite, direction) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(metabolite) %>%
    mutate(total = sum(count)) %>%
    ungroup() %>%
    mutate(metabolite = factor(metabolite, levels = rev(unique(metabolite[order(-total)]))))
}

side_mimics_counts    <- make_side_bar(fgsea_mimics_sig)
side_opposites_counts <- make_side_bar(fgsea_opposites_sig)
```

## plot — mimics

```{r}
ggplot(side_mimics_counts, aes(x = metabolite, y = count, fill = direction)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Positive" = "#D82632", "Negative" = "#2166AC")) +
  labs(x = "Metabolite", y = "Number of pathways", fill = "NES direction") +
  paper_theme +
  theme(axis.text.y = element_text(size = 12)) +
  coord_flip() +
  scale_y_reverse()
```

## plot — opposites

```{r}
ggplot(side_opposites_counts, aes(x = metabolite, y = count, fill = direction)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Positive" = "#D82632", "Negative" = "#2166AC")) +
  labs(x = "Metabolite", y = "Number of pathways", fill = "NES direction") +
  paper_theme +
  theme(axis.text.y = element_text(size = 12)) +
  coord_flip() +
  scale_y_reverse()
```

## save reference tables

```{r}
write.csv(side_mimics_counts,
          "~/Documents/gutmicrobiome_pharmacology/supp_fig1c_side_bar_table.csv", row.names = FALSE)
write.csv(side_opposites_counts,
          "~/Documents/gutmicrobiome_pharmacology/supp_fig2c_side_bar_table.csv", row.names = FALSE)
```

# update 4-8-2026

## dot plot

```{r}
fgsea_mimics    <- read.csv("~/Documents/gutmicrobiome_pharmacology/fgsea_top_mimics_regorafenib_metabolites-4-2-2026.csv")
fgsea_opposites <- read.csv("~/Documents/gutmicrobiome_pharmacology/fgsea_top_opposites_regorafenib_metabolites-4-2-2026.csv")

fgsea_mimics_sig <- fgsea_mimics %>%
  filter(padj < 0.05) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway   = clean_pathway(pathway)
  )

fgsea_opposites_sig <- fgsea_opposites %>%
  filter(padj < 0.05) %>%
  mutate(
    direction = ifelse(NES > 0, "Positive", "Negative"),
    pathway   = clean_pathway(pathway)
  )

fgsea_mimics_sig_strict    <- fgsea_mimics_sig %>% filter(padj < 0.01)
fgsea_opposites_sig_strict <- fgsea_opposites_sig %>% filter(padj < 0.01)
```

## for re-running

```{r}
fgsea_mimics_sig_strict    <- read.csv("~/Documents/gutmicrobiome_pharmacology/sfig1a_dotplot_fgsea.csv")
fgsea_opposites_sig_strict <- read.csv("~/Documents/gutmicrobiome_pharmacology/sfig2a_dotplot_fgsea.csv")

#head(unique(fgsea_mimics_sig_strict$metabolite))

# Supp Fig 1 - mimics dot plot (padj < 0.01)
p_supp1 <- ggplot(fgsea_mimics_sig_strict, aes(x = metabolite, y = pathway,
                              size = -log10(padj),
                              color = NES)) +
  geom_point() +
  scale_color_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#D55E00",
    midpoint = 0
  ) +
  labs(
    x     = "Metabolite",
    y     = "Pathway",
    size  = "-log10(FDR)",
    color = "NES"
  ) +
  paper_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13)
  )

# Supp Fig 2 - opposites dot plot (padj < 0.01)
p_supp2 <- ggplot(fgsea_opposites_sig_strict, aes(x = metabolite, y = pathway,
                                 size = -log10(padj),
                                 color = NES)) +
  geom_point() +
  scale_color_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#D55E00",
    midpoint = 0
  ) +
  labs(
    x     = "Metabolite",
    y     = "Pathway",
    size  = "-log10(FDR)",
    color = "NES"
  ) +
  paper_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13)
  )

# --- Supp Fig 1: mimics ---
n_path_s1  <- length(unique(fgsea_mimics_sig_strict$pathway))
n_metab_s1 <- length(unique(fgsea_mimics_sig_strict$metabolite))
max_label_chars_s1 <- max(nchar(unique(fgsea_mimics_sig_strict$metabolite)), na.rm = TRUE)
label_margin_s1    <- max_label_chars_s1 * 0.04  # ~0.04in per character at 45 degrees
width_s1 <- 9.8
height_s1 <- 8

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig1_mimics_dotplot.svg",
       plot = p_supp1, device = svglite::svglite,
       width = width_s1, height = height_s1, units = "in")

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig1_mimics_dotplot.pdf",
       plot = p_supp1, device = "pdf",
       width = width_s1, height = height_s1, units = "in")

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig1_mimics_dotplot.tiff",
       plot = p_supp1, device = "tiff",
       width = width_s1, height = height_s1, units = "in", dpi = 300)

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig1_mimics_dotplot.png",
       plot = p_supp1, device = "png",
       width = width_s1, height = height_s1, units = "in", dpi = 300)

# --- Supp Fig 2: opposites ---
n_path_s2  <- length(unique(fgsea_opposites_sig_strict$pathway))
n_metab_s2 <- length(unique(fgsea_opposites_sig_strict$metabolite))
width_s2   <- 6.5
height_s2  <- max(5,  1.5 + n_path_s2  * 0.3)

max_label_chars_s2 <- max(nchar(unique(fgsea_opposites_sig_strict$metabolite)), na.rm = TRUE)
label_margin_s2    <- max_label_chars_s2 * 0.04

height_s2 <- max(5, 1.5 + n_path_s2 * 0.3 + label_margin_s2)

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig2_opposites_dotplot.svg",
       plot = p_supp2, device = svglite::svglite,
       width = width_s2, height = height_s2, units = "in")

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig2_opposites_dotplot.pdf",
       plot = p_supp2, device = "pdf",
       width = width_s2, height = height_s2, units = "in")

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig2_opposites_dotplot.tiff",
       plot = p_supp2, device = "tiff",
       width = width_s2, height = height_s2, units = "in", dpi = 300)

ggsave("~/Documents/gutmicrobiome_pharmacology/suppfig2_opposites_dotplot.png",
       plot = p_supp2, device = "png",
       width = width_s2, height = height_s2, units = "in", dpi = 300)
```

```{r}
fgsea_mimics_sig_strict %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ", ")) %>%
  write.csv("~/Documents/gutmicrobiome_pharmacology/sfig1a_dotplot_fgsea.csv",
            row.names = FALSE)

fgsea_opposites_sig_strict %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ", ")) %>%
  write.csv("~/Documents/gutmicrobiome_pharmacology/sfig2a_dotplot_fgsea.csv",
            row.names = FALSE)
```

# update supplementary table 1

## mimics 

```{r}
# fix double-formatted names when reading in
fix_double_format <- function(x) {
  # "metabolite (method (id))" -> "metabolite (method_id)"
  gsub("\\(([^)]+) \\(([^)]+)\\)\\)", "(\\1_\\2)", x)
}

supp_table1_long <- read.csv("~/Documents/gutmicrobiome_pharmacology/supp_fig1a_intersection_pathway_names.csv") %>%
  mutate(metabolites_in_intersection = fix_double_format(metabolites_in_intersection)) %>%
  # expand pathways to one per row first
  separate_rows(pathways, sep = "; ") %>%
  mutate(pathways = clean_pathway(pathways)) %>%
  dplyr::rename(
    "Pathway"               = pathways,
    "Metabolites"           = metabolites_in_intersection,
    "Number of metabolites" = num_metabolites_in_intersection,
    "Overlap type"          = overlap_type
  ) %>%
  dplyr::select(
    "Pathway",
    "Number of metabolites",
    "Metabolites",
    "Overlap type"
  ) %>%
  arrange(desc(`Number of metabolites`), Pathway)

write.csv(supp_table1_long,
          "~/Documents/gutmicrobiome_pharmacology/Supplementary_Table1a_mimics_pathways.csv",
          row.names = FALSE)

# check dimensions
nrow(supp_table1_long)
head(supp_table1_long %>% dplyr::select(Pathway, `Number of metabolites`))
```

## opposites

```{r}
supp_table2_long <- read.csv("~/Documents/gutmicrobiome_pharmacology/supp_fig2a_intersection_pathway_names.csv") %>%
  mutate(metabolites_in_intersection = fix_double_format(metabolites_in_intersection)) %>%
  separate_rows(pathways, sep = "; ") %>%
  mutate(pathways = clean_pathway(pathways)) %>%
  dplyr::rename(
    "Pathway"               = pathways,
    "Metabolites"           = metabolites_in_intersection,
    "Number of metabolites" = num_metabolites_in_intersection,
    "Overlap type"          = overlap_type
  ) %>%
  dplyr::select(
    "Pathway",
    "Number of metabolites",
    "Metabolites",
    "Overlap type"
  ) %>%
  arrange(desc(`Number of metabolites`), Pathway)

write.csv(supp_table2_long,
          "~/Documents/gutmicrobiome_pharmacology/Supplementary_Table1b_opposites_pathways.csv",
          row.names = FALSE)

nrow(supp_table2_long)
head(supp_table2_long %>% dplyr::select(Pathway, `Number of metabolites`))
```

# update supplementary table 2

```{r}
shared_pathway_directions <- fgsea_mimics_sig %>%
  filter(pathway %in% shared_pathways) %>%
  group_by(pathway) %>%
  summarise(
    mean_NES_concordant      = round(mean(NES), 3),
    n_metabolites_concordant = n()
  ) %>%
  inner_join(                          # changed from left_join to inner_join
    fgsea_opposites_sig %>%
      filter(pathway %in% shared_pathways) %>%
      group_by(pathway) %>%
      summarise(
        mean_NES_discordant      = round(mean(NES), 3),
        n_metabolites_discordant = n()
      ),
    by = "pathway"
  ) %>%
  mutate(
    direction_concordant = ifelse(mean_NES_concordant > 0, "Positive", "Negative"),
    direction_discordant = ifelse(mean_NES_discordant > 0, "Positive", "Negative"),
    opposite_directions  = direction_concordant != direction_discordant
  ) %>%
  arrange(desc(abs(mean_NES_concordant))) %>%
  dplyr::rename(
    "Pathway"                    = pathway,
    "Mean NES (concordant)"      = mean_NES_concordant,
    "Mean NES (discordant)"      = mean_NES_discordant,
    "N metabolites (concordant)" = n_metabolites_concordant,
    "N metabolites (discordant)" = n_metabolites_discordant,
    "NES direction (concordant)" = direction_concordant,
    "NES direction (discordant)" = direction_discordant,
    "Opposite directions"        = opposite_directions
  )

cat("Total shared pathways:", nrow(shared_pathway_directions), "\n")
cat("Pathways with opposite directions:", 
    sum(shared_pathway_directions$`Opposite directions`), "\n")

write.csv(shared_pathway_directions,
          "~/Documents/gutmicrobiome_pharmacology/Supplementary_Table2_shared_pathway_directions.csv",
          row.names = FALSE)
```